VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-15 doc/publications/journal2014/paper.tex
update table

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}. % 80
% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
Our Hadoop implementation performed the worst with 14mins and 23secs. % 863
The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\label{fivewaycomparison}
\begin{table}[h!]
\caption{The resources consumption of the five compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
      \begin{tabular}{|c|ccc|ccc|c|c|}
        \hline
           Tool &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering} & Accuracy \\
& threads & memory & time  &threads & memory & time  & \\
  \hline
\variantSpark{}	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& 0.84	\\ 
ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& x	\\
Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& x	\\
R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& x	\\
Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\
  \hline
      \end{tabular}
\end{table}

>>>>

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}. % 80
% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
% AIDAN: They were all run on Azure. I was just generating ARI results with Python the other day on another machine
We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
Our Hadoop implementation performed the worst with 14mins and 23secs. % 863
The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\label{fivewaycomparison}
\begin{table}[h!]
\caption{The resources consumption of the five compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
      \begin{tabular}{|c|ccc|ccc|c|c|}
        \hline
           Tool &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering} & Accuracy \\
& threads & memory & time  &threads & memory & time  & \\
  \hline
\variantSpark{}	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& 0.84	\\ 
ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& 0.84	\\
Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& 0.84	\\
R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& 0.84	\\
Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\
ADMIXTURE	& 1	& 32	& 10min 8sec	& 8	& 32 & 8min 19sec	& 0.25	\\
  \hline
      \end{tabular}
\end{table}
